Model Recognition as Planning

### 2018-09-21 doc/icaps-19/model-recognition.tex
paper icaps19: pages [1-5] stable. Ready for review

The task of {\em model recognition} can also be understood as a classification task where each class is represented with a different planning model and the observed plan execution is the example to classify. In this case, the planning model that is associated to a class is acting as a class prototype that the summarizes all the plan executions that could be synthesized with that model or in other words, all the examples that belong to that class.

@@

@@

\section{Model recognition with classical planning}
This section shows that, when the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, then $\delta(\mathcal{M},\tau)$ can be computed with a compilation of a classical planning with conditional effects. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces an optimal solution plan $\pi^*_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi^*_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan with a prefix (steps [0,1]) for editing a given {\em blockswold} model where the positive effects {\tt\small (handempty)} and {\tt\small (clear ?v1)} of the {\tt\small stack} schema are missing. The postfix of the plan (steps [2,6]) validates the edited action model at the observation of a four action plan for inverting a two-block tower where intermediate states, $s_1$, $s_2$ and $s_3$, are unobserved. 
\begin{figure}
{\tt\scriptsize
00 : (insert\_add\_handempty\_stack)\\
01 : (insert\_add\_clear\_stack\_var1)\\
02 : (apply\_unstack blockB blockA i1 i2)\\
03 : (apply\_putdown blockB i2 i3)\\
04 : (apply\_pickup blockA i3 i4)\\
05 : (apply\_stack blockA blockB i4 i5)\\
06 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-1]) and validating (steps [2-6]) a given \strips\ planning model for the {\em blockswold}.}
\label{fig:plan-pdistance}
\end{figure}

Note that our interest is not in $\mathcal{M}'$, the edited model resulting from the compilation, but in the number of required {\em edit operations} (insertions and deletions) required by $\mathcal{M}'$ to be validated. In the example of Figure~\ref{fig:plan-pdistance} $\delta(\mathcal{M},\tau)=2$ and $\delta(\mathcal{M},*)=3\times 2\times (11+5)$ since there are 4 action schemes ({\small\tt pickup}, {\small\tt putdown}, {\small\tt stack} and {\small\tt unstack}) s.t. $|F_v|=|F_{\tt stack}|=|F_{\tt unstack}|=11$ while $|F_{\tt pickup}|=|F_{\tt putdown}|=5$.

@@

\subsection{A propositional encoding for \strips\ action schema}
We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

Given a \strips\ action schema $\xi$, a propositional encoding for the {\em preconditions}, {\em negative} and {\em positive} effects of that schema can be represented with fluents $[pre|del|add]\_p\_name(\xi)$. Figure~\ref{fig:encoding} shows the propositional encoding for the six action schema previously defined in Figure~\ref{fig:model-example}. The interest of having a propositional encoding for \strips\ action schema is that it allows to define {\em editable actions} that is, actions whose semantics is given by the value of these particular fluents on the current state.

Formally given an operator schema $\xi\in\mathcal{M}$ we can build its {\em editable} version as follows:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
\cond(\mathsf{editable_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}.
\end{align*}
\end{small}
bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:editable} shows the PDDL encoding of the action for applying a programmed operator. For instance, Figure~\ref{fig:editable} shows the editable version of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (see Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} holds, behaves exactly as is defined in Figure~\ref{fig:model-example}. Further it allows to swap the semantics of two programable schemas provided that they are {\em comparable}.

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action inc-x
  :parameters (?v1 ?v2)
  :precondition
    (and (or (not (pre_xcoord_v1_inc-x)) (xcoord ?v1))
         (or (not (pre_xcoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_ycoord_v1_inc-x)) (xcoord ?v1))                       
         (or (not (pre_ycoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_q0__inc-x)) (q0))
         (or (not (pre_q1__inc-x)) (q1)))
         (or (not (pre_next_v1_v1_inc-x)) (next ?v1 ?v1)))
         (or (not (pre_next_v1_v2_inc-x)) (next ?v1 ?v2)))
         (or (not (pre_next_v2_v1_inc-x)) (next ?v2 ?v1)))
         (or (not (pre_next_v2_v2_inc-x)) (next ?v2 ?v2))))
    :effect (and
       (when (del_xcoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_xcoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_ycoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_ycoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_q0__inc-x) (not (q0)))
       (when (del_q1__inc-x) (not (q1)))
       (when (del_next_v1_v1_inc-x) (not (next ?v1 ?v1)))
       (when (del_next_v1_v2_inc-x) (not (next ?v1 ?v2)))
       (when (del_next_v2_v1_inc-x) (not (next ?v2 ?v1)))
       (when (del_next_v2_v2_inc-x) (not (next ?v2 ?v2)))

>>>>

The task of {\em model recognition} can be understood as a classification task where each class is represented with a different planning model and the observed plan execution is the example to classify. In this case, the planning model that is associated to a class is acting as a class prototype that the summarizes all the plan executions that could be synthesized with that model or in other words, all the examples that belong to that class.

@@

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four opertator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

\section{Model recognition with classical planning}
This section shows that, for \strips\ planning models, the $P(\tau|\mathcal{M})$ likelihood can be estimated with a classical planning compilation. In more detail, when the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, then $\delta(\mathcal{M},\tau)$ can be computed with a compilation of a classical planning with conditional effects.

The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$}. The solution plan continues with a postfix that:
\begin{enumerate}
\item Induces an optimal solution plan $\pi^*_\tau$ for the original classical planning problem $P$.
\item Validates that $\tau$ is an observation of the execution of $\pi^*_\tau$ on the classical planning problem $P$.
\end{enumerate}
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan with a prefix (steps [0,6]) for editing the planning model of Figure\ref{fig:model-example} when its action {\tt\small inc-x} is defined without preconditions and its positive/negative effects are swapped wrt Figure~\ref{fig:model-example}. The postfix of the plan (steps [7,13]) validates the edited action model at the observation of the plan execution illustrated at Figure~\ref{fig:grid-example}. 
\begin{figure}
{\tt\scriptsize
00 : (insert\_pre\_xcoord\_v1\_inc-x)\\
01 : (insert\_pre\_next\_v1\_v2\_inc-x)\\
02 : (insert\_pre\_q0\_inc-x)\\
03 : (delete\_del\_xcoord\_v2\_inc-x)\\
04 : (delete\_add\_xcoord\_v1\_inc-x)\\
05 : (insert\_del\_xcoord\_v1\_inc-x)\\
06 : (insert\_add\_xcoord\_v2\_inc-x)\\
07 : (editable\_inc-x 1 2)\\
08 : (editable\_inc-x 2 3)\\
09 : (editable\_inc-x 3 4)\\
10 : (editable\_inc-x 4 5)\\
11 : (editable\_inc-y-even 1 2)\\
12 : (editable\_dec-x 5 4)\\
13 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-6]) and validating (steps [7-13]) the planning model of Figure\ref{fig:model-example} when action {\tt\small inc-x} has no preconditions and positive/negative effects are swapped wrt Figure\ref{fig:model-example}.}
\label{fig:plan-pdistance}
\end{figure}

Note that our interest is not in $\mathcal{M}'$, the edited model resulting from the compilation, but in the number of required {\em edit operations} (insertions and deletions) required by $\mathcal{M}'$ to be validated. In the example of Figure~\ref{fig:plan-pdistance} $\delta(\mathcal{M},\tau)=7$ and $\delta(\mathcal{M},*)=6\times 3\times 10$ since there are 6 action schemes for which $|{\mathcal I}_{\Psi,\xi}|=10$.

@@

\subsection{A propositional encoding for \strips\ action schema}
Given a \strips\ action schema $\xi$, a propositional encoding for the {\em preconditions}, {\em negative} and {\em positive} effects of that schema can be represented with fluents of the kind $[pre|del|add]\_e\_name(\xi)$ such that $e\in{\mathcal I}_{\Psi,\xi}$ is a single element from the set of interpretations of predicates $\Psi$ over the corresponding objects $\Omega_\xi$. Figure~\ref{fig:encoding} shows the propositional encoding for the six action schema previously defined in Figure~\ref{fig:model-example}.

The interest of having a propositional encoding for \strips\ action schema is that it allows to define {\em editable actions} that is, actions whose semantics is given by the value of these particular fluents on the current state. Formally given an operator schema $\xi\in\mathcal{M}$ its {\em editable} version can be compactly specified with {\em conditional effects}:
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{editable_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\\
\cond(\mathsf{editable_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}.
\end{align*}
\end{small}
Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega\subseteq\Omega^{ar(\xi)}$ that appear at the same position. Figure~\ref{fig:editable} shows the PDDL encoding of the {\em editable version} of the {\tt\small inc-x(?v1,?v2)} schema for robot navigation in a $n\times n$ grid (see Figure~\ref{fig:model-example}). Note that this editable schema, when the fluents of Figure~\ref{fig:encoding} hold, behaves exactly as defined in Figure~\ref{fig:model-example}. Further, it allows to swap the semantics of any two schema provided that they are {\em comparable}.

\begin{figure}
  \begin{tiny}  
  \begin{verbatim}
(:action editable_inc-x
  :parameters (?v1 ?v2)
  :precondition
    (and (or (not (pre_xcoord_v1_inc-x)) (xcoord ?v1))
         (or (not (pre_xcoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_ycoord_v1_inc-x)) (xcoord ?v1))                       
         (or (not (pre_ycoord_v2_inc-x)) (xcoord ?v2))
         (or (not (pre_q0__inc-x)) (q0))
         (or (not (pre_q1__inc-x)) (q1)))
         (or (not (pre_next_v1_v1_inc-x)) (next ?v1 ?v1)))
         (or (not (pre_next_v1_v2_inc-x)) (next ?v1 ?v2)))
         (or (not (pre_next_v2_v1_inc-x)) (next ?v2 ?v1)))
         (or (not (pre_next_v2_v2_inc-x)) (next ?v2 ?v2))))
    :effect (and
       (when (del_xcoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_xcoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_ycoord_v1_inc-x) (not (xcoord ?v1)))
       (when (del_ycoord_v2_inc-x) (not (xcoord ?v2)))
       (when (del_q0__inc-x) (not (q0)))
       (when (del_q1__inc-x) (not (q1)))
       (when (del_next_v1_v1_inc-x) (not (next ?v1 ?v1)))
       (when (del_next_v1_v2_inc-x) (not (next ?v1 ?v2)))
       (when (del_next_v2_v1_inc-x) (not (next ?v2 ?v1)))
       (when (del_next_v2_v2_inc-x) (not (next ?v2 ?v2)))
